Longitudinal Manifold Estimation

### 2023-02-09 paper/main.tex
minor change to test compilation
updates to introduction

Manifold learning is an approach to modeling high-dimensional data that is assumed to lie along a low-dimensional manifold. These approaches have become increasingly popular for medical imaging applications, where assuming an underlying manifold structure is helpful for modeling certain anatomical structures. In these settings, interacting with a structure in a low-dimensional form may be more intuitive than in a higher-dimensional space. For example, Yue et al. (2016) seeks to compute a parametric representation of the corpus callosum. A number of methods have been developed for manifold learning, including Isomap (Tenenbaum et al. 2000), locally linear embedding (Roweis and Saul, 2000), and Laplacian eigenmaps (Belkin and Niyogi, 2003). Meng and Eloyan (2021) proposed a principal manifold framework that resolved several limitations of previously developed extensions of the principal curves approach introduced by Hastie and Stuetzle (1989).

@@

[DISCUSS LONGITUDINAL DIMENSION REDUCTION APPROACHES CURRENTLY AVAILABLE - NEED TO REVIEW ARTICLES]

The remainder of this article is laid out as follows: In Section 2, we introduce the principal manifold estimation method proposed in Meng and Eloyan (2021), discuss how this approach can be adapted for longitudinal settings, and show the algorithm being used to accomplish this. Section 3 demonstrates the performance of this approach on simulated data. In Section 4, we apply this method to longitudinal data of the hippocampus in individuals with Alzheimer's disease. The article finishes with a discussion of the method in Section 5.

\section{Methods}

@@

Questions:
\begin{enumerate}
  \item What is the best way to evaluate the performance of the estimates? We can use SSD for the embedded values, but how to assess the parameterized values?
  \item Should there be updates to what is defined as the "optimal" model in the PME algorithm? Currently it is possible for a model with non-minimal SSD to be chosen within the fitting of a single tuning value.
  \item Why would we see the SSD increase substantially in late optimization steps?
  \item Are there cases where the SSD and MSD differ substantially? Why would that be the case?
  \item Are there any techniques that help with large amounts of noise in the data? Data reduction should help with outliers, but likely not consistently high noise levels. Early indications that possibly allowing for higher tuning parameters would help?
  \item Why does PME algorithm struggle on Swiss Roll cross-section? It seems to work well up until spiral reaches second layer. The manifold should never intersect, but from the angle from the origin does overlap.
  \item Generally, what are the limitations on the estimated manifolds intersecting? How close can it come to intersecting without showing strange behavior?
  \item Is thin-plate spline computational strategy from Wahba the best approach, or could updated computational approach help (Wood, 2003)?
\end{enumerate}

>>>>

Manifold learning describes a set of statistical approaches to modeling high-dimensional data that are assumed to lie along a low-dimensional manifold. These methods have become increasingly popular for medical imaging applications, where assuming an underlying manifold structure is helpful for modeling certain anatomical shapes. In these settings, interacting with a structure in a low-dimensional form may be more intuitive than in a higher-dimensional space. For example, Yue et al. (2016) seeks to compute a parametric representation of the corpus callosum. A number of methods have been developed for manifold learning, including Isomap (Tenenbaum et al. 2000), locally linear embedding (Roweis and Saul, 2000), and Laplacian eigenmaps (Belkin and Niyogi, 2003). Meng and Eloyan (2021) proposed a principal manifold framework that resolved several limitations of previously developed extensions of the principal curves approach introduced by Hastie and Stuetzle (1989).

@@

Some work has been conducted on longitudinal dimension reduction methods, but this research has primarily focused on linear methods, such as the longitudinal principal component analysis approach proposed by Kinson et al. (2019). To our knowledge, the approach to longitudinal smoothing in the manifold learning setting taken in this paper is novel.

The remainder of this article is laid out as follows: In Section 2, we introduce the principal manifold estimation method proposed in Meng and Eloyan (2021). Section 3 discusses how this approach can be adapted for longitudinal settings, and shows the algorithm being used to accomplish this. Section 4 demonstrates the performance of this approach on simulated data. In Section 5, we apply this method to longitudinal data of the hippocampus in individuals with Alzheimer's disease. The article finishes with a discussion of the method in Section 6.

\section{Principal Manifolds and Estimation Algorithm}

Because the algorithm proposed in this article relies on the same definitions and notation as those introduced in Meng and Eloyan (2021), we provide it here. To generalize the principal curve framework proposed in Hastie and Stuetzle (1989) to intrinsic dimensions $d > 1$, Meng and Eloyan introduce the function space 
\[%
  C_{\infty}(\mathbb{R}^d \to \mathbb{R}^D) = \left\{f \in C()\right\}
.\]%

\section{Longitudinal Principal Manifold Estimation Algorithm}

@@

Questions:
\begin{enumerate}
  \item What is the best way to evaluate the performance of the estimates? We can use SSD for the embedded values, but how to assess the parameterized values?
  \item Should there be updates to what is defined as the "optimal" model in the PME algorithm? Currently it is possible for a model with non-minimal SSD to be chosen within the fitting of a single tuning value.
  \item Why would we see the SSD increase substantially in late optimization steps?
  \item Are there cases where the SSD and MSD differ substantially? Why would that be the case?
  \item Are there any techniques that help with large amounts of noise in the data? Data reduction should help with outliers, but likely not consistently high noise levels. Early indications that possibly allowing for higher tuning parameters would help?
  \item Why does PME algorithm struggle on Swiss Roll cross-section? It seems to work well up until spiral reaches second layer. The manifold should never intersect, but from the angle from the origin does overlap.
  \item Generally, what are the limitations on the estimated manifolds intersecting? How close can it come to intersecting without showing strange behavior?
  \item Is thin-plate spline computational strategy from Wahba the best approach, or could updated computational approach help (Wood, 2003)?
  \item How much error does the registration process introduce? Can the LPME algorithm successfully handle a certain number of bad registrations or extractions?
\end{enumerate}
